|

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add "orthogonality" assumption slides

>>>>

## Orthogonality {.smaller}

:::::: columns
::: column
```{r item-oblique-axes-1}
#| fig-align:  center
#| fig-height: 8
#| fig-width:  8
mirt_item <- tribble(
  ~item, ~a_1, ~a_2, ~d,
  1,     0.5,  1,    0
)

mirt_item_coords |>
  ggplot(
    aes(
      origin_1, origin_2,
      xend = end_1,
      yend = end_2,
      color = factor(item), fill = factor(item)
    ),
  )                                                  +
  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
  geom_segment(
    data = item_spikes,
    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
    linewidth = LINE_WIDTH,
    linetype  = '43',
    color     = PALETTE_BLUE["light"]
  ) +
  geom_segment(
    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin  = "mitre",
    linewidth = VECTOR_WIDTH
  ) +
  scale_x_continuous(
    limits       = c(-0.75, 1.5),
    breaks       = (-1:3)/2,
    minor_breaks = NULL,
    name         = NULL,
    expand       = expansion(),
    sec.axis     = dup_axis(name = "*&theta;*^*^~2~", labels = NULL)
  ) +
  scale_y_continuous(
    limits       = c(-0.75, 1.5),
    breaks       = (-1:2)/2,
    minor_breaks = NULL,
    name         = NULL,
    expand       = expansion(),
    sec.axis     = dup_axis(name = "*&theta;*^*^~1~", labels = NULL)
  ) +
  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
  annotate(
    "richtext",
    x = c(0.5, -0.1), y = c(-0.1, 1),
    label         = c("*a*^*^~11~", "*a*^*^~21~"),
    size          = 8,
    label.colour  = NA,
    label.padding = unit(0, "lines"),
    fill = NA
  ) +
  coord_fixed(clip = "on") +
  theme(
    axis.line          = element_blank(),
    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
    axis.title.x       = element_markdown(hjust = 1/3),
    axis.title.y.right = element_markdown(vjust = 1/3),
    axis.ticks         = element_blank(),
    panel.grid.major   = element_line(
      linewidth = GRID_WIDTH,
      linetype  = "17"
    )
  )
```
:::

:::: column
$$
  \begin{align}
    \mathbf{a}^*_1 &=
      \begin{bmatrix}
        0.5 \\
        1
      \end{bmatrix}\\
      d^*_1 &= 0
  \end{align}
$$

::: fragment
$$
  \begin{align}
    MDISC^*_1 &= \sqrt{
        [0.5, 1]
        \begin{bmatrix}
          0.5 \\
          1
        \end{bmatrix}
      } \\
      &= \sqrt{ 0.5 · 0.5 + 1 · 1 } \\
      &= \sqrt{ 0.25 + 1 } \\
      &= \sqrt{ 1.25 } \approx 1.12
  \end{align}
$$
:::
::::
::::::

::: notes
You'll recall that we needed orthogonal axes to apply Reckase's formulas.

So let's say we have these "theta star" axes,...

...where we have this item, with discrimination (.5, 1) in this coordinate system, where we may compute its MDISC.

We can transform these axes to another coordinate system, for example, applying an oblique rotation...
:::

## Orthogonality {.smaller}

::::: columns
::: column
```{r item-oblique-axes-2}
#| fig-align:  center
#| fig-height: 8
#| fig-width:  8
BASIS_VEC_1 <- c(1, 0)
BASIS_VEC_2 <- c(-.5, 1)

::: {.column .fragment}
<br> <br>

--- @reckase_difficulty_1985 [p. 404]
:::
:::::

::: notes
...to obtain these "theta axes" (let's say, because they align with personality traits of our interest).
We have now different discrimination components than in the "theta star" axes.

If we applied here Reckase's restriction...
:::

## Orthogonality {.smaller}

::::: columns
::: column
```{r item-oblique-axes-4}
#| fig-align:  center
#| fig-height: 8
#| fig-width:  8
mirt_item_transf <- tribble(
  ~item, ~a_1, ~a_2, ~d,
  1,     1,    1,    0
)

mirt_item_params_transf <- mirt_item_transf |>
  compute_mirt_params(d, starts_with('a'))

mirt_item_coords_transf <- mirt_item_params_transf |>
  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
    rename_with(str_remove, pattern = "transf_")

item_spikes <- mirt_item_coords_transf |>
  slice(1 |> rep(2))                   |>
  add_column(dim = 'F' |> paste0(1:2)) |>
  mutate(
    origin_1 = if_else(dim == 'F1', end_1, origin_1),
    origin_2 = if_else(dim == 'F2', end_2, origin_2),
  )

mirt_item_coords_transf |>
  ggplot(
    aes(
      origin_1, origin_2,
      xend = end_1,
      yend = end_2,
      color = factor(item), fill = factor(item)
    ),
  )                                                  +
  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
  geom_segment(
    data = item_spikes,
    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
    linewidth = LINE_WIDTH,
    linetype  = '43',
    color     = PALETTE_BLUE["light"]
  ) +
  geom_segment(
    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin  = "mitre",
    linewidth = VECTOR_WIDTH
  ) +
  scale_x_continuous(
    limits       = c(-0.75, 1.5),
    breaks       = (-1:3)/2,
    minor_breaks = NULL,
    name         = NULL,
    expand       = expansion(),
    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
  ) +
  scale_y_continuous(
    limits       = c(-0.75, 1.5),
    breaks       = (-1:2)/2,
    minor_breaks = NULL,
    name         = NULL,
    expand       = expansion(),
    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
  ) +
  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
  annotate(
    "richtext",
    x = c(1, -0.1), y = c(-0.1, 1),
    label         = c("*a*~11~", "*a*~21~"),
    size          = 8,
    label.colour  = NA,
    label.padding = unit(0, "lines"),
    fill = NA
  ) +
  coord_fixed(clip = "on") +
  theme(
    axis.line          = element_blank(),
    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
    axis.title.x       = element_markdown(hjust = 1/3),
    axis.title.y.right = element_markdown(vjust = 1/3),
    axis.ticks         = element_blank(),
    panel.grid.major   = element_line(
      linewidth = GRID_WIDTH,
      linetype  = "17"
    )
  )
```
:::

::: column
$$
  \begin{align}
    \mathbf{a}_1 &=
      \begin{bmatrix}
        1 \\
        1
      \end{bmatrix}\\
      d_1 &= 0
  \end{align}
$$

$$
  \begin{align}
    MDISC_1 &= \sqrt{
        [1, 1]
        \begin{bmatrix}
          1 \\
          1
        \end{bmatrix}
      } \\
      &= \sqrt{ 1 · 1 + 1 · 1 } \\
      &= \sqrt{ 1 + 1 } \\
      &= \sqrt{ 2 } \approx 1.41
  \end{align}
$$
:::
:::::

::: notes
...we would be assuming that the axes are actually orthogonal, implying that the MDISC is not invariant.

As you can see, does not seem intuitively right to assume that both coordinate systems are orthogonal, ¿right?
:::

## Orthogonality {.smaller}

::::: columns
::: column
```{r item-oblique-axes-2}
```
:::

::: column
<br> <br>

--- @reckase_difficulty_1985 [p. 404]
:::
:::::

::: notes
What we actually need is to compute the multidimensional parameters in this general space, which is not necessarily orthogonal (or, we should say more generally, not necessarily orthonormal).
:::
